---
title: "inspercidados Package Testing"
subtitle: "Comprehensive functionality tests"
author: "inspercidados"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
editor: visual
---

# Setup

```{r setup, error=TRUE}
library(inspercidados)

# Real DOIs from dataverse.insper.edu.br
dois <- c(
  geoses = "10.60873/FK2/7IXFPX",           # GeoSES
  ilhas_calor = "10.60873/FK2/NQA7LY",      # Ilhas de Calor
  pop_censo = "10.60873/FK2/GTO7DD"         # População e domicílios (censo)
)

cat("✓ Package loaded successfully\n")
cat("Testing with", length(dois), "real DOIs from Insper Dataverse\n")
```

# 1. List Available Datasets

## 1.1 List aliases only

```{r test-list-aliases, error=TRUE}
cat("=== Testing list_available_datasets() ===\n\n")

aliases <- list_available_datasets()
print(aliases)
cat("\nFound", nrow(aliases), "datasets with aliases\n")
```

## 1.2 List with Dataverse search

```{r test-list-search, error=TRUE}
cat("=== Testing Dataverse search ===\n\n")

# Search for datasets
search_results <- list_available_datasets(
  search = "GeoSES",
  include_search = TRUE
)
print(search_results)
```

## 1.3 Search by collection

```{r test-list-collection, error=TRUE}
cat("=== Testing collection search ===\n\n")

# Try to search a collection (may need to adjust collection name)
collection_results <- list_available_datasets(
  collection = "insper-cidades",
  include_search = TRUE
)
print(collection_results)
```

# 2. Identifier Resolution

## 2.1 Test DOI detection

```{r test-doi-detection, error=TRUE}
cat("=== Testing DOI detection ===\n\n")

test_strings <- c(
  "10.60873/FK2/7IXFPX",
  "doi:10.60873/FK2/7IXFPX",
  "iptu_sp",
  "not-a-doi"
)

for (s in test_strings) {
  result <- inspercidados:::is_doi(s)
  cat(sprintf("is_doi('%s'): %s\n", s, result))
}
```

## 2.2 Test identifier resolution

```{r test-resolve-identifier, error=TRUE}
cat("=== Testing identifier resolution ===\n\n")

# Test with DOI
cat("1. Resolving DOI:\n")
resolved_doi <- inspercidados:::resolve_identifier(dois["geoses"])
cat("  Input:", dois["geoses"], "\n")
cat("  DOI:", resolved_doi$doi, "\n")
cat("  Source:", resolved_doi$source, "\n\n")

# Test with alias
cat("2. Resolving alias:\n")
resolved_alias <- inspercidados:::resolve_identifier("iptu_sp")
cat("  Input: iptu_sp\n")
cat("  DOI:", resolved_alias$doi, "\n")
cat("  Source:", resolved_alias$source, "\n\n")

# Test with invalid identifier (should error)
cat("3. Resolving invalid identifier (should error):\n")
tryCatch({
  inspercidados:::resolve_identifier("invalid_dataset_xyz")
}, error = function(e) {
  cat("  Error (expected):", e$message, "\n")
})
```

# 3. List Files in Datasets

## 3.1 List files using DOI

```{r test-list-files-doi, error=TRUE}
cat("=== Testing list_files() with DOI ===\n\n")

for (name in names(dois)) {
  cat("\nDataset:", name, "\n")
  cat("DOI:", dois[name], "\n")

  tryCatch({
    files <- list_files(dois[name])
    print(files)
  }, error = function(e) {
    cat("Error:", e$message, "\n")
  })
}
```

## 3.2 List files using alias

```{r test-list-files-alias, error=TRUE}
cat("=== Testing list_files() with alias ===\n\n")

tryCatch({
  files_alias <- list_files("iptu_sp")
  print(files_alias)
}, error = function(e) {
  cat("Error:", e$message, "\n")
})
```

# 4. Get Dataset Information

## 4.1 Get dataset info with DOI

```{r test-get-info, error=TRUE}
cat("=== Testing get_dataset_info() ===\n\n")

for (name in names(dois)) {
  cat("\n", strrep("=", 60), "\n")
  cat("Dataset:", name, "\n")
  cat(strrep("=", 60), "\n")

  tryCatch({
    info <- get_dataset_info(dois[name])
    cat("DOI:", info$doi, "\n")
    cat("Server:", info$server, "\n")
    cat("Dataverse version:", info$dataverse_version, "\n")
    cat("Publication date:", info$publication_date, "\n")

    if (!is.null(info$files)) {
      cat("\nFiles:\n")
      print(head(info$files, 3))
    }
  }, error = function(e) {
    cat("Error:", e$message, "\n")
  })
}
```

# 5. Download Datasets

## 5.1 Download using DOI (first dataset)

```{r test-download-doi, error=TRUE}
cat("=== Testing get_dataset() with DOI ===\n\n")

cat("Attempting to download GeoSES dataset...\n")
tryCatch({
  data_geoses <- get_dataset(dois["geoses"])

  cat("\n✓ Dataset loaded successfully!\n")
  cat("Dimensions:", nrow(data_geoses), "rows ×", ncol(data_geoses), "columns\n")
  cat("\nFirst few column names:\n")
  print(head(names(data_geoses), 10))

  cat("\nDataset attributes:\n")
  cat("  Source:", attr(data_geoses, "source"), "\n")
  cat("  Dataset:", attr(data_geoses, "dataset"), "\n")
  cat("  DOI:", attr(data_geoses, "doi"), "\n")
  cat("  Filename:", attr(data_geoses, "filename"), "\n")
  cat("  Download date:", attr(data_geoses, "download_date"), "\n")

}, error = function(e) {
  cat("Error downloading dataset:", e$message, "\n")
})
```

## 5.2 Download using alias

```{r test-download-alias, error=TRUE}
cat("=== Testing get_dataset() with alias ===\n\n")

tryCatch({
  data_alias <- get_dataset("iptu_sp")

  cat("\n✓ Dataset loaded successfully!\n")
  cat("Dimensions:", nrow(data_alias), "rows ×", ncol(data_alias), "columns\n")

}, error = function(e) {
  cat("Error:", e$message, "\n")
})
```

## 5.3 Download with specific filename

```{r test-download-filename, error=TRUE}
cat("=== Testing get_dataset() with specific filename ===\n\n")

# First, list files to see what's available
tryCatch({
  files <- list_files(dois["pop_censo"])
  cat("Available files:\n")
  print(files)

  if (nrow(files) > 0) {
    target_file <- files$filename[1]
    cat("\nAttempting to download file:", target_file, "\n")

    data_specific <- get_dataset(dois["pop_censo"], filename = target_file)
    cat("✓ Downloaded successfully!\n")
    cat("Dimensions:", nrow(data_specific), "rows ×", ncol(data_specific), "columns\n")
  }
}, error = function(e) {
  cat("Error:", e$message, "\n")
})
```

## 5.4 Test force download

```{r test-force-download, error=TRUE}
cat("=== Testing force_download parameter ===\n\n")

tryCatch({
  # Download with force (should re-download even if cached)
  data_force <- get_dataset(dois["geoses"], force_download = TRUE)
  cat("✓ Force download completed\n")
  cat("Dimensions:", nrow(data_force), "rows ×", ncol(data_force), "columns\n")
}, error = function(e) {
  cat("Error:", e$message, "\n")
})
```

# 6. Citations

## 6.1 Text citation with DOI

```{r test-citation-text, error=TRUE}
cat("=== Testing cite_dataset() - text format ===\n\n")

for (name in names(dois)) {
  cat("\nDataset:", name, "\n")
  cat(strrep("-", 60), "\n")

  tryCatch({
    citation <- cite_dataset(dois[name], format = "text")
    cat(citation, "\n")
  }, error = function(e) {
    cat("Error:", e$message, "\n")
  })
}
```

## 6.2 BibTeX citation

```{r test-citation-bibtex, error=TRUE}
cat("=== Testing cite_dataset() - bibtex format ===\n\n")

tryCatch({
  bibtex <- cite_dataset(dois["geoses"], format = "bibtex")
  cat(bibtex, "\n")
}, error = function(e) {
  cat("Error:", e$message, "\n")
})
```

## 6.3 RIS citation

```{r test-citation-ris, error=TRUE}
cat("=== Testing cite_dataset() - RIS format ===\n\n")

tryCatch({
  ris <- cite_dataset(dois["geoses"], format = "ris")
  cat(ris, "\n")
}, error = function(e) {
  cat("Error:", e$message, "\n")
})
```

## 6.4 Print citation

```{r test-print-citation, error=TRUE}
cat("=== Testing print_citation() ===\n\n")

tryCatch({
  print_citation(dois["geoses"])
}, error = function(e) {
  cat("Error:", e$message, "\n")
})
```

## 6.5 Package citation

```{r test-package-citation, error=TRUE}
cat("=== Testing cite_package() ===\n\n")

cat("Text format:\n")
cite_package("text")

cat("\n\nBibTeX format:\n")
cite_package("bibtex")
```

# 7. Cache Management

## 7.1 List cached datasets

```{r test-list-cache, error=TRUE}
cat("=== Testing list_cached_datasets() ===\n\n")

cached <- list_cached_datasets()
print(cached)
cat("\nTotal cached datasets:", nrow(cached), "\n")
cat("Total cache size:", sum(cached$size_mb, na.rm = TRUE), "MB\n")
```

## 7.2 Get cache directory

```{r test-cache-dir, error=TRUE}
cat("=== Testing get_cache_dir() ===\n\n")

cache_dir <- inspercidados:::get_cache_dir()
cat("Cache directory:", cache_dir, "\n")
cat("Exists:", dir.exists(cache_dir), "\n")

if (dir.exists(cache_dir)) {
  files <- list.files(cache_dir, pattern = "\\.parquet$")
  cat("Files in cache:", length(files), "\n")
}
```

## 7.3 Clear specific dataset cache (commented out to preserve downloads)

```{r test-clear-cache, error=TRUE, eval=FALSE}
cat("=== Testing clear_cache() ===\n\n")

# Clear a specific dataset (uncomment to test)
# clear_cache("geoses")

# Clear all cache (uncomment to test)
# clear_cache()

cat("Cache clearing tests skipped to preserve downloads.\n")
cat("Uncomment code above to test cache clearing.\n")
```

# 8. Edge Cases and Error Handling

## 8.1 Invalid DOI

```{r test-invalid-doi, error=TRUE}
cat("=== Testing with invalid DOI ===\n\n")

tryCatch({
  data_invalid <- get_dataset("10.1234/INVALID/DOI")
}, error = function(e) {
  cat("Error (expected):", e$message, "\n")
})
```

## 8.2 Non-existent alias

```{r test-invalid-alias, error=TRUE}
cat("=== Testing with non-existent alias ===\n\n")

tryCatch({
  data_invalid <- get_dataset("nonexistent_dataset_xyz")
}, error = function(e) {
  cat("Error (expected):", e$message, "\n")
})
```

## 8.3 Invalid citation format

```{r test-invalid-format, error=TRUE}
cat("=== Testing with invalid citation format ===\n\n")

tryCatch({
  citation <- cite_dataset(dois["geoses"], format = "invalid_format")
}, error = function(e) {
  cat("Error (expected):", e$message, "\n")
})
```

# 9. Show Citation Option

## 9.1 Download with citation display

```{r test-show-citation, error=TRUE}
cat("=== Testing show_citation parameter ===\n\n")

tryCatch({
  data_with_citation <- get_dataset(dois["geoses"], show_citation = TRUE)
  cat("\n✓ Dataset loaded with citation displayed\n")
}, error = function(e) {
  cat("Error:", e$message, "\n")
})
```

# 10. Summary

```{r test-summary, error=TRUE}
cat("=== Testing Summary ===\n\n")

cat("Package version:", as.character(packageVersion("inspercidados")), "\n")
cat("R version:", R.version.string, "\n")
cat("\n")

cat("Available functions:\n")
cat("  - list_available_datasets()\n")
cat("  - list_files()\n")
cat("  - get_dataset()\n")
cat("  - get_dataset_info()\n")
cat("  - cite_dataset()\n")
cat("  - print_citation()\n")
cat("  - cite_package()\n")
cat("  - list_cached_datasets()\n")
cat("  - clear_cache()\n")
cat("\n")

cat("Tested with", length(dois), "real DOIs:\n")
for (name in names(dois)) {
  cat("  -", name, ":", dois[name], "\n")
}

cat("\n✓ All tests completed!\n")
```

# Session Info

```{r session-info, error=TRUE}
sessionInfo()
```
