---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

`inspercidados` provides access to curated Brazilian urban datasets produced by
[Insper Cidades](https://www.insper.edu.br) and hosted on Insper's Dataverse.
The package is a thin wrapper around the
[`dataverse`](https://CRAN.R-project.org/package=dataverse) package.

## Installation

```{r install}
# Install from GitHub (development version)
remotes::install_github("portalcidados/inspercidados")
```

## Discovering datasets

`list_datasets()` returns a tibble of all available datasets.
The search is local — no network call is made.

```{r list}
library(inspercidados)

# All datasets
ds <- list_datasets()
ds

# Search by keyword, theme, or region
ds <- list_datasets("Mobilidade")
ds

# Use dplyr or base R to filter further
subset(ds, grepl("Mobilidade", theme, ignore.case = TRUE))
```

The tibble has columns `alias`, `title`, `description`, `theme`, `region`,
`keywords`, and `doi`.
Use the `alias` column as the identifier in all other functions.

## Downloading a dataset

`get_dataset()` downloads data directly into R.
It accepts an **alias**, a bare **DOI**, or a full **DOI URL**.

```{r download-alias}
# Easiest: use the alias
embarques <- get_dataset("embarques_mensais")
embarques
```

```{r download-doi}
# Equivalent: use the DOI
embarques <- get_dataset("10.60873/FK2/BPYHFB")
```

```{r download-url}
# Or the full DOI URL
embarques <- get_dataset("https://doi.org/10.60873/FK2/BPYHFB")
```

When a dataset contains multiple files you can narrow down the selection:

```{r download-filters}
# Filter by year (matches year anywhere in the filename)
iptu_2024 <- get_dataset("iptu_sp", year = 2024)

# Request a specific file by exact name
iptu_geo <- get_dataset("iptu_sp", filename = "iptu_2024.gpkg")

# Use a regex pattern
iptu_geo <- get_dataset("iptu_sp", file_pattern = "\\.gpkg$")
```

## Downloading with documentation

By default `get_dataset()` returns only the data.
Set `docs = TRUE` to also retrieve the variable documentation.
When an Excel documentation file (`documentacao_*.xlsx`) is present in the
dataset, it is downloaded automatically; otherwise Dataverse metadata is
returned instead.

```{r download-docs}
result <- get_dataset("embarques_mensais", docs = TRUE)

# The dataset
result$data

# The documentation (tibble from xlsx, or list of metadata fields)
result$docs
```

## Spatial datasets

Datasets that contain spatial data (geometries) require the `sf` package.
The package will warn you if `sf` is not loaded when you request a spatial
dataset.

```{r spatial}
library(sf)

alvaras <- get_dataset("alvaras_sp", file_pattern = "\\.gpkg$")
plot(sf::st_geometry(alvaras))
```

## Citing a dataset

`cite_dataset()` fetches author and publication metadata from Dataverse and
formats a citation. Three formats are supported.

```{r cite}
# Plain text (default)
cite_dataset("embarques_mensais")

# BibTeX — paste into your .bib file
cite_dataset("embarques_mensais", format = "bibtex")

# RIS — import into Zotero, Mendeley, or EndNote
cite_dataset("embarques_mensais", format = "ris")
```

## Getting the companion R script

Some datasets come with companion R scripts. `get_script()` downloads a script
and opens it in your editor. Two types are available:

- **`"analysis"`** (default): a getting-started script demonstrating common
  analyses with the dataset.
- **`"replication"`**: the production pipeline that generated the dataset.
  Provided for transparency and reproducibility; requires internal resources
  and cannot be run by external users.

```{r script}
# Opens the analysis script in your editor (default)
get_script("embarques_mensais")

# Opens the replication script (how the data was produced)
get_script("embarques_mensais", type = "replication")

# Download only (do not open)
path <- get_script("embarques_mensais", open = FALSE)
source(path)
```
